Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-10-03 ms/main.tex
Fix corollaries 4 & 6 and reorder them

With only oblique and horizontal transmission, i.e. $v = 0$, Eq.\ \ref{eq:horizontal} becomes $\hat{p}=p$ and Eq.\ \ref{eq:nextgen_adults_slimpify} becomes % TODO check this, I changed the ref from nextgen_parents_simplified to nextgen_adults_slimpify
\begin{equation}  \label{eq:nextgen_parents_oblique_only}
p' = p^2 (T_B-T_A) + p (1+T_A-T_B) \;,
\end{equation}

@@

\begin{corollary}[Intermediate association of transmission and cooperation]\label{corollary:intermediate_association}
Cooperation will increase from rarity if social association is high enough, specifically if
  \begin{equation} \label{eq:unequal_transmission_from_rarity_in_alpha_terms}
    \begin{aligned}
    T_A > T_B &\;\; \text{and} \;\; a_2 < \alpha, \quad \text{or} \\
    T_A < T_B &\;\; \text{and} \;\; a_1 < \alpha \,.
    \end{aligned}
    \end{equation} 
\end{corollary}

\autoref{fig:result2}c demonstrates these conditions.
With horizontal transmission bias for cooperation ($T_A>T_B$), cooperation can fix from any initial frequency if $a_2<\alpha$ (green area in positive x-axis). 
Otherwise, cooperation can also fix if the horizontal bias is strong, the initial frequency is high, and social association is high, $a_1 < \alpha$ (blue area).
With horizontal bias favoring defection ($T_A<T_B$), cooperation can fix from any frequency if social association is high, $a_1<\alpha$ (green area with $T_A<T_B$), and can also increase when rare and reach stable coexistence with defection if social association is intermediate, $a_2<\alpha$ (yellow area).
Without horizontal bias ($T_A=T_B$) fixation of cooperation occurs if social association is high enough; $\frac{c}{b} \cdot \frac{1-T}{T} < \alpha$ (inequality~\ref{eq:equal_transmission}; in this case $a_1=a_2$).

@@

In general, this condition cannot be formulated in the form of Hamilton's rule due to the horizontal transmission bias $T_A-T_B$.
Without horizontal transmission bias, i.e., with $T=T_A=T_B$, these conditions reduce to a form of Hamilton's rule.
\\
\begin{corollary}[Necessary condition for fixation of cooperation]
\label{corollary:vhat}
  Fixation of cooperation is possible only if the vertical transmission rate is high enough,
  \begin{equation} \label{eq:fixation_of_cooperation_vertical_transmission_condition}
    \begin{aligned}
      v>\hat{v} = \frac{T_B - T_A}{1-T_A} \;.
    \end{aligned}
    \end{equation} 
\end{corollary}
%Note that $\sigma \frac{\partial \gamma_1}{\partial v} = \sigma (T_B-T_A)$ and $\sigma \frac{\partial \hat\gamma_2}{\partial v} = \sigma (T_A-T_B)(b \alpha T_B - (1+b)(1-T_A))$
Note that this condition does not guarantee fixation of cooperation. Rather, if this condition does not apply then cooperation cannot fix.
When horizontal transmission is biased for cooperation, $T_A>T_B$, cooperation can fix with any vertical transmission rate, i.e. $\hat{v}<0$.
In contrast, if the horizontal transmission is biased for defection, $T_A<T_B$,  cooperation can fix only if the vertical transmission rate is high enough: in this case oblique transmission can prevent fixation of cooperation, see \autoref{fig:result3}bd.
\\

In Corollary~\autoref{corollary:intermediate_association}, we examined the effect of social association on evolution of cooperation in the case of perfect vertical transmission ($v=1$).
A more general case can be stated for $0 < v \leq 1$.
Consider the social association boundaries 
\begin{equation}\begin{aligned}\label{eq:boundries_assortative_meeting_general_case}
  \hat{a}_1 = \frac{c\cdot v(1-T_A) -(T_A-T_B)(1+b-c)}{b\cdot v \cdot T_B}, \quad
  \hat{a}_2 = \frac{c\cdot v(1-T_B)-(T_A-T_B)}{b\cdot v\cdot T_A} \;.
\end{aligned}\end{equation}
Note that when $T_A>T_B$ then $\partial \hat a_2 / \partial v = b T_A (T_A-T_B) > 0$, and when $T_A<T_B$ then $\partial \hat a_1 / \partial v = b(1+b-c)T_B(T_A-T_B) < 0$.
Similarly, $\partial \hat a_1 / \partial T_A < 0$, $\partial \hat a_2 / \partial T_A < 0$, $\partial \hat a_2 / \partial T_B > 0$, and $\text{sign}(\partial \hat a_1 / \partial T_B) = \text{sign}\big(T_A (1+b-c) - (1-T_A)cv\big)$.
%TODO remove derivatives of a1

@@

\begin{corollary}[Intermediate association of transmission and cooperation]
\label{corollary:intermediate_association_res3}
Cooperation will increase from rarity if social association is high enough, specifically if
  \begin{equation} \label{eq:unequal_transmission_from_rarity_in_alpha_terms}
    \begin{aligned}
    T_A > T_B &\;\; \text{and} \;\; \hat a_2 < \alpha, \quad \text{or} \\
    T_A < T_B &\;\; \text{and} \;\; \hat a_1 < \alpha \,.
    \end{aligned}
    \end{equation} 
\end{corollary}
\autoref{fig:result3}cd illustrates this conditions.
%TODO fix for increasing from rarity, only using a2

Note that increasing the horizontal transmission of cooperation, $T_A$, decreases both boundaries ($\partial \hat a_1 / \partial T_A, \partial \hat a_2 / \partial T_A < 0$), thus facilitating evolution of cooperation. 
Increasing the horizontal transmission of defection, increases one boundary, ($\partial \hat a_2 / \partial T_B > 0$).
So, if there is horizontal bias for cooperation ($T_A>T_B$), decreasing that bias will make it harder for cooperation to evolve.

Interestingly, when there is horizontal transmission bias for cooperation ($T_A>T_B$), then vertical transmission hinders the evolution of cooperation ($\partial \hat a_2 / \partial v > 0$). In contrast, when there is horizontal transmission bias for defection ($T_A<T_B$), vertical transmission facilitates the evolution of cooperation ($\partial \hat a_1 / \partial v < 0$).
\\
%TODO switch corollary 7 and 6

@@

%%% Figure: Spatial model
\begin{figure}[h]
  \centering
    \includegraphics[scale=0.5]{spatial1.png}
    ~
    \includegraphics[scale=0.5]{spatial2.png} \\
  \caption{
  \textbf{Evolution of cooperation in a spatial model.}
  The expected frequency of cooperators in a structured population after 10,000 generations is shown (red for 0\%, green for 100\%) as function of both the cost of cooperation ($c$) on the y-axis, and the symmetric horizontal transmission rate ($T=T_A=T_B$) on the x-axis of the left panel, or the transmission bias $T_A-T_B$ on the x-axis of the right panel.
  The population evolves on a 100-by-100 grid. Cooperation and horizontal cultural transmission are all local between adjacent sites, and each site had 8 neighbors.
  The black curves represent the cost boundaries for the evolution of cooperation in a well-mixed population with social association where $\alpha=1/8$ in (left) \autoref{eq:equal_transmission} and (right) \label{eq:cost_boundaries}.
  Simulations were stopped at generation 10,000 or if one of the phenotypes fixed and 50 simulations were executed for each parameter set.
  Here, population size is $10,000$ (100-by-100 grid), benefit of cooperation, $b=1.3$, perfect vertical transmission $v=1$.
  \textbf{(left)} Symmetric horizontal transmission, $T=T_A=T_B$. \textbf{(right)} Horizontal transmission rates $T_A=0.4$, $0.3<T_B<0.5$.
  }
  \label{fig:spatial}
\end{figure}

>>>>

With only oblique and horizontal transmission, i.e. $v = 0$, Eq.\ \ref{eq:horizontal} becomes $\hat{p}=p$ and Eq.\ \ref{eq:nextgen_adults_slimpify} becomes 
\begin{equation}  \label{eq:nextgen_parents_oblique_only}
p' = p^2 (T_B-T_A) + p (1+T_A-T_B) \;,
\end{equation}

@@

\begin{corollary}[Intermediate association of transmission and cooperation]\label{corollary:intermediate_association}
Cooperation will increase from rarity if social association is high enough, specifically if
  \begin{equation} \label{eq:unequal_transmission_from_rarity_in_alpha_terms}
    \begin{aligned}
     a_2 < \alpha
    \end{aligned}
    \end{equation} 
\end{corollary}

\autoref{fig:result2}c demonstrates these conditions.
With horizontal transmission bias for cooperation ($T_A>T_B$), cooperation can fix from any initial frequency if $a_2<\alpha$ (green area in positive x-axis). 
With horizontal bias favoring defection ($T_A<T_B$), cooperation can fix from any frequency if social association is high, $a_1<\alpha$ (green area with $T_A<T_B$), and can also increase when rare and reach stable coexistence with defection if social association is intermediate, $a_2<\alpha$ (yellow area).
Without horizontal bias ($T_A=T_B$) fixation of cooperation occurs if social association is high enough; $\frac{c}{b} \cdot \frac{1-T}{T} < \alpha$ (inequality~\ref{eq:equal_transmission}; in this case $a_1=a_2$).

@@

In general, this condition cannot be formulated in the form of Hamilton's rule due to the horizontal transmission bias $T_A-T_B$.
Without horizontal transmission bias, i.e., with $T=T_A=T_B$, these conditions reduce to a form of Hamilton's rule.
\\

In Corollary~\autoref{corollary:intermediate_association}, we examined the effect of social association on evolution of cooperation in the case of perfect vertical transmission ($v=1$).
A more general case can be stated for $0 < v \leq 1$.
Consider the social association boundaries 
\begin{equation}\begin{aligned}\label{eq:boundries_assortative_meeting_general_case}
  \hat{a}_1 = \frac{c\cdot v(1-T_A) -(T_A-T_B)(1+b-c)}{b\cdot v \cdot T_B}, \quad
  \hat{a}_2 = \frac{c\cdot v(1-T_B)-(T_A-T_B)}{b\cdot v\cdot T_A} \;.
\end{aligned}\end{equation}

@@

\begin{corollary}[Intermediate association of transmission and cooperation]
\label{corollary:intermediate_association_res3}
Cooperation will increase from rarity if social association is high enough, specifically if
  \begin{equation} \label{eq:unequal_transmission_from_rarity_in_alpha_terms}
    \begin{aligned}
     \hat a_2 < \alpha.
    \end{aligned}
    \end{equation} 
\end{corollary}
\autoref{fig:result3}cd illustrates this conditions.

Note that when $T_A>T_B$ then $\partial \hat a_2 / \partial v = b T_A (T_A-T_B) > 0$, and when $T_A<T_B$ then $\partial \hat a_1 / \partial v = b(1+b-c)T_B(T_A-T_B) < 0$. When $T_A>T_B$ and the vertical transmission $v$ increases the boundary $\hat{a}_2$ also increases
and therefore, a greater value of social association is required for fixation of cooperation from rarity. When $T_B>T_A$ decreasing of $v$ increases the boundary $\hat{a}_2$ 
and therefore, a greater value of social association is required for fixation of cooperation from rarity.
\\

\begin{corollary}[Necessary condition for fixation of cooperation]
\label{corollary:vhat}
  Fixation of cooperation is possible only if the vertical transmission rate is high enough,
  \begin{equation} \label{eq:fixation_of_cooperation_vertical_transmission_condition}
    \begin{aligned}
      v>\hat{v} = \frac{T_B - T_A}{1-T_A} \;.
    \end{aligned}
    \end{equation} 
\end{corollary}
%Note that $\sigma \frac{\partial \gamma_1}{\partial v} = \sigma (T_B-T_A)$ and $\sigma \frac{\partial \hat\gamma_2}{\partial v} = \sigma (T_A-T_B)(b \alpha T_B - (1+b)(1-T_A))$
Note that this condition does not guarantee fixation of cooperation. Rather, if this condition does not apply then cooperation cannot fix.
When horizontal transmission is biased for cooperation, $T_A>T_B$, cooperation can fix with any vertical transmission rate, i.e. $\hat{v}<0$.
In contrast, if the horizontal transmission is biased for defection, $T_A<T_B$,  cooperation can fix only if the vertical transmission rate is high enough: in this case oblique transmission can prevent fixation of cooperation, see \autoref{fig:result3}bd.
\\

@@

%%% Figure: Spatial model
\begin{figure}[h]
  \centering
%   \includegraphics[scale=0.5]{spatial1.png}
    ~
 %   \includegraphics[scale=0.5]{spatial2.png} \\
  \caption{
  \textbf{Evolution of cooperation in a spatial model.}
  The expected frequency of cooperators in a structured population after 10,000 generations is shown (red for 0\%, green for 100\%) as function of both the cost of cooperation ($c$) on the y-axis, and the symmetric horizontal transmission rate ($T=T_A=T_B$) on the x-axis of the left panel, or the transmission bias $T_A-T_B$ on the x-axis of the right panel.
  The population evolves on a 100-by-100 grid. Cooperation and horizontal cultural transmission are all local between adjacent sites, and each site had 8 neighbors.
  The black curves represent the cost boundaries for the evolution of cooperation in a well-mixed population with social association where $\alpha=1/8$ in (left) \autoref{eq:equal_transmission} and (right) \label{eq:cost_boundaries}.
  Simulations were stopped at generation 10,000 or if one of the phenotypes fixed and 50 simulations were executed for each parameter set.
  Here, population size is $10,000$ (100-by-100 grid), benefit of cooperation, $b=1.3$, perfect vertical transmission $v=1$.
  \textbf{(left)} Symmetric horizontal transmission, $T=T_A=T_B$. \textbf{(right)} Horizontal transmission rates $T_A=0.4$, $0.3<T_B<0.5$.
  }
  \label{fig:spatial}
\end{figure}
